---
title: Characterising South Africa’s unregulated drug markets - findings from a cross-sectional survey of people who supply sell heroin, methamphetamine and cocaine 
author:
  - name: MJ Stowe
    orcid: 0000-0002-0760-5497
    corresponding: true
    email: mjstowe@kirby.unsw.edu.au
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - The Kirby Institute, UNSW Syndey, Sydney Australia
      - The South African Network of People Who Use Drugs (SANPUD), South Africa
      - Community Oriented Primary Care Research Unit, Department of Family Medicine, University of Pretoria, South Africa  
  - name: Shaun Shelly
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The South African Network of People Who Use Drugs (SANPUD), South Africa
      - Community Oriented Primary Care Research Unit, Department of Family Medicine, University of Pretoria, South Africa
      - TB HIV Care, South Africa 
  - name: Jason Eligh
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - The Global Initiative Against Transnational Organised Crime (GI-TOC)
  - name: Andrew Scheibe
    orcid: 0000-0002-7859-8394
    corresponding: false
    roles: []
    affiliations:
      - TB HIV Care, South Africa

keywords:
  - People who supply drugs
  - Drug markets
  - Heroin
  - Cocaine
  - Methamphetamine
  
  
abstract: |                                                                                
  **Background:** The value and effects of southern Africa’s drug market is increasing. However, little is known about drug sellers and the nature of drug sales in the region. We examined the characteristics of people who supply heroin, methamphetamine and cocaine (drugs) at retail and wholesale levels in South Africa. **Methods:** Participants in this cross-sectional study included people who sell drugs who were recruited from nine South African cities between May – November 2020. Participants completed a questionnaire assessing seller attributes and sale characteristics. Data was analysed using descriptive statistics. **Results:**Among the 176 participants, people sold heroin (n=139), methamphetamine (n=62) and cocaine (n=85), with X% selling  trnah oend[[sjv	. Overall, 93% (n=159) were male, X% sold multiple drugs  and people had been working for a median of 5 years (IQR   X – X). Overall, X% (n=X) sold to consumers (retail), X% (n=X) to sellers and consumers (wholesale ), and X% (n=X) exclusively to sellers. Among retail sellers, the median weekly amount of heroin, methamphetamine and cocaine sold was Xg (IQR X – X), Xg (IQR X – X), and Xg (IQR X – X), respectively, to X and X and X consumers, respectively. Among wholesale sellers, heroin, methamphetamine and cocaine was sold at a median of   Xg (IQR Xg – Xg), Xg (IQR Xg – Xg), and Xg (IQR Xg – Xg) to X and X and X sellers, respectively.**Conclusion:**This study is the first to describe the characteristics of people who sell drugs and nature of sales in South Africa. These data could be useful to inform definitions and thresholds for possession and supply in drug legislation.            

plain-language-summary: |
  To be written. 
date: last-modified
css: custom.css
# bibliography: references.bib
citation:
  container-title: South African Drug Market
numbered-sections: true
---


<!-- ## ABSTRACT -->

<!-- ### BACKGROUND -->

<!-- The value and effects of southern Africa’s drug market is increasing. However, little is known about drug sellers and the nature of drug sales in the region. We examined the characteristics of people who supply heroin, methamphetamine and cocaine (drugs) at retail and wholesale levels in South Africa. -->

<!-- ### METHODS -->

<!-- Participants in this cross-sectional study included people who sell drugs who were recruited from nine South African cities between May – November 2020. Participants completed a questionnaire assessing seller attributes and sale characteristics. Data was analysed using descriptive statistics.  -->

<!-- ### RESULTS -->

<!-- Among the 176 participants, people sold heroin (n=139), methamphetamine (n=62) and cocaine (n=85), with X% selling  trnah oend[[sjv	. Overall, 93% (n=159) were male, X% sold multiple drugs  and people had been working for a median of 5 years (IQR   X – X). Overall, X% (n=X) sold to consumers (retail), X% (n=X) to sellers and consumers (wholesale ), and X% (n=X) exclusively to sellers. Among retail sellers, the median weekly amount of heroin, methamphetamine and cocaine sold was Xg (IQR   X – X), Xg (IQR   X – X), and Xg (IQR   X – X), respectively, to X and X and X consumers, respectively. Among wholesale sellers, heroin, methamphetamine and cocaine was sold at a median of   Xg (IQR   Xg – Xg), Xg (IQR   Xg – Xg), and Xg (IQR   Xg – Xg) to X and X and X sellers,   respectively.  -->

<!-- ### CONCLUSION -->

<!-- This study is the first to describe the characteristics of people who sell drugs and nature of sales in South Africa. These data could be useful to inform definitions and thresholds for possession and supply in drug legislation.  -->


## Introduction

## Data & Methods {#sec-data-methods}

### Study design and participants {#sec-study-design}

In this cross-sectional study, participants were enrolled from  a range of sites including open air drug markets, private dwellings, areas that are serviced by harm reduction outreach services and discrete private locations. There was national representation of sites across all states and territories in Australia (except Tasmania), including Australian Capital Territory
(n=3), New South Wales (n=37), the Northern Territory (n=1), Queensland
(n=9), South Australia (n=6), Victoria (n=2), and Western Australia
(n=1). Study enrolment took place between October 2020 and April 2021,
and continued throughout periods of COVID-19-related restrictions. To be eligible for this study, participants had to be aged 18 years or older,
provide voluntary and informed consent, use opioids regularly. In this
study, regular opioid use was defined as use of any opioid (including
heroin, methadone, burpenorphine use and other opioids or the
extramedical use of pharmaceutical opioids) on at least 21 of the past
28 days. The content within this article is presented following the
'Strengthening the reporting of observational studies in epidemiology'
(STROBE) guidelines [@von2007; @Vandenbroucke2007]. A completed
checklist of the STROBE requirements is available in the
@sec-supp-material.

### Procedures {#sec-procedures}

Fieldworkers

Training

Participants were recruited from a range of settings (including drug and alcohol treatment services, needle and syringe programmes, organisations providing support and services for people who use drugs, community-based and -led advocacy organisations, and pharmacies), via snowballing (where
eligible participants promote the study via their personal networks),
and word-of-mouth. The study was advertised at services using posters
and fliers, but service staff were not directly involved in recruitment.
Participants who were interested in the study contacted the study team
directly and were screened for eligibility over the telephone.

Interviewers contacted the person over the phone or video conference
(Microsoft Teams or Zoom) to obtain consent and conduct the interview.
Interviews were semi-structured were conducted by trained interviewers.
Participants completed an interviewer-administered questionnaire focused
on patient preferences for OAT that included information on participant
demographics, drug use characteristics, and drug treatment. The
interviews took approximately 30-45 minutes to complete, and
participants were reimbursed AUD\$40 for their time and out-of-pocket
expenses.

### Outcomes {#sec-outcomes}

The primary outcome was a preference for opioid agonist therapy with
methadone, which was assessed by asking the question "*Of all the
following types of medications used for opioid agonist therapy, if you
could choose today, which one would you prefer?*" Options included
methadone, monobuprenorphine (Subutex), buprenorphine-naloxone
(Suboxone) taken orally as a film or tablet, long acting injectable
buprenorphine (Buvidal or Sublocade), any treatment (i.e. no preference
for any medication), and no treatment.

Demographic, behavioural, and clinically significant factors
hypothesised to be associated with preference to receive OAT with
methadone were determined a priori, comprising the following: 1) age at
survey (18 - 35; 36 -- 45; \> 45 years old) ), 2) gender (female; male;
transgender), 3) education (\< year 10; \> year 10), 4) employment
(paid; other), 5) homelessness (no; yes), 6) chronic pain (no; yes), 7)
incarceration history (never, \>6 months ago, \\6 months ago), 8) Drugs
used and injected in the last 28 days (heroin, non-prescribed methadone,
non-prescribed buprenorphine, other pharmaceutical opioids,
methamphetamine, cocaine and benzodiazepines), 9) Frequency of injection
drug use (\>1 year ago, within 1--12 months ago, within the last month
less than daily, and daily or more), 10) OAT (history and current), with
methadone (history and current) and buprenorphine (history and current).
Among the sub-group of participants currently receiving OAT, factors
hypothesized to be associated with a preference for OAT with methadone
included 11) current OAT medication (methadone; buprenorphine;
buprenorphine-naloxone; long-acting injectable buprenorphine; none), 12)
Drugs used while on OAT (no, heroin, non-prescribed methadone,
non-prescribed buprenorphine, other pharmaceutical opioids,
methamphetamine, cocaine and benzodiazepines); 13) Pay out of pocket for
oat treatment (no; yes); 14) Amount paid out of pocket for OAT (xx; xx),
15) Site last prescribed OAT medication was collected (pharmacy; public
clinic; private clinic; at home; other), 16) Frequency of OAT dose
collection (daily or several times per week; weekly or less frequently),
17) Distance traveled to OAT collection site, 18) Time traveled to OAT
collection site (less than 30 minutes; more than 30 minutes).

### Statistical analysis {#sec-analysis}

The  of people who reported preference for methadone and
factors associated with this preference was assessed. Quantitative
parameters are presented as the median and interquartile range \[IQR\].
Categorical parameters are presented as the number and percentage (n;
%). The association between the response to each variable and the
binarized category of preference for methadone was explored using
logistic regression modeling through unadjusted bivariate analysis,
providing an odds ratio and the 95 % confidence interval (OR \[95%
CI\]), and comparisons adjusted for age category and gender, providing
an adjusted OR (aOR) and the 95 % CI. A sample of those who would choose
treatment and who had a preference for methadone or any buprenorphine
forumation was selected for analysis (n=352) . Multivariable logistic
regression was used assess factors associated with preferred medication
for OAT. Unadjusted and adjusted odds ratios were derived using logistic
regression and 95% confidence intervals calculated for all variables.
For the final model, all the explanatory variables significantly
associated with the outcome in the bivariate models were included in the
multivariable model, whilst also adjusting for age and gender. The final
model was assessed through examining which contributed significantly to
model fit through backwards selection using decreasing values of Akaike
Information Criterion. The variance inflation factor was used to assess
collinearity in the final model [@obrien2007]. Individuals with missing
values were not integrated in the models. A subset of the sample
comprising of participants currently receiving OAT with methadone and
buprenorphine (including sublingual-monobuprenorphine,
buprenorphine-naloxone and long-acting injectable buprenorphine) was
selected for analysis. There was no formal sample size calculation,
however, based of similar previous surveys (e.g. [@larance2020;
@rolland2021]) we aimed to recruit 300 - 400 people.

The analytical approach received input from and was reviewed by people
who use drugs and by people with lived- and living-experience of drug selling. Advice and perspectives were provided on the inclusion of
explanatory variables and the clinical and practical significance of
identified factors.

The analysis was undertaken in using `R` version
`r paste0(version$major, "." ,version$minor)` using the following
packages: gtsummary v. 1.7.2 [@gtsummary], finalfit v. 1.0.6
[@finalfit], quarto v. 1.3 [@quarto] and rmarkdown v. 2.25 [@rmarkdown].
The code for this analysis is available on request from the
corresponding author. The data supporting the findings of this study are
available within the article and its supporting information. The
analysis was not pre-registered and the results should be considered
exploratory.

### Study oversight {#sec-oversight}

All participants provided written informed consent before study
procedures. The study protocol were approved by the Human Research
Ethics Committees at St Vincent's Hospital, Sydney (HREC Ref:
HREC/17/SVH/113) and the Aboriginal Health and Medical Research Council
(HREC Ref: 1279/17). This study was conducted according to the
Declaration of Helsinki and International Conference on Harmonization
Good Clinical Practice (ICH/GCP) guidelines.

### Role of the Funding source {#sec-funding}

The study was funded by a research grant from X. The funders had no role
in the study design, data collection, analysis, interpretation of the
results, writing or the decision to submit the study for publication.
JG, AC, LD, and MJS had access to the raw data. The National Drug and
Alcohol Research Centre and the Kirby Institute, UNSW Sydney
collaborated to design the study, monitor study conduct, and perform the
statistical analysis. X, X and X were responsible for the decision to
submit the study for publication.

## RESULTS

```{r}
# Drug Market Tables
# Tables

# Load packages
pacman::p_load(
  ggplot2 ,    # plotting and graphing results
  rlang,      # read in excel files
  forcats,    # read in excel files
  broom,      # tidy code
  readxl,     # read in excel files
  pacman,     # loading and reading in packages
  rio,        # importing data
  here,       # relative file pathways
  janitor,    # data cleaning and tables
  lubridate,  # working with dates
  matchmaker, # dictionary-based cleaning
  epikit,     # age_categories() function
  tidyverse,  # data management and visualization
  gtsummary,  # logistic regression and plotting results
  gt,         # for gtsummary
  flextable,  # table creation and manipulating
  finalfit,   # logistic regression and plotting results
  jtools,
  corrplot,
  codebookr,
  codebook,
  sjlabelled,
  likert,
  kableExtra,
  haven,
  tableone, 
  summarytools
)

# load data
#
load(
  here::here("data_processed", "drug_market_working.rda"))

load(
  here::here("data_processed", "drug_market_factors.rda"))

# Create a theme for figures, tables and all processed text

apa_theme <- function (ft)  {
  ft %>%
    flextable::font(fontname = "Times New Roman", part = "all") %>%
    flextable::fontsize(size = 10, part = "all") %>%
    flextable::align(align = "left", part = "all") %>%
    flextable::align(align = "left", part = "header") %>%
    #flextable::rotate(rotation = "lrtb", align = "top", part = "body") %>%
    flextable::border_remove() %>%
    flextable::hline_top(border = officer::fp_border(width = 1.5), part = "all") %>%
    flextable::hline_bottom(border = officer::fp_border(width = 1.5), part = "all") %>%
    flextable::hline_bottom(border = officer::fp_border(width = 1.5), part = "header") %>%
    flextable::hline_bottom(border = officer::fp_border(width = 1.5), part = "footer") %>%
    flextable::hline_bottom(border = officer::fp_border(width = 2), part = "all") %>%
    flextable::autofit()
}

# Set the theme to compact [reduce spacing around text and numbers]

## Set the theme as a compact [spacing]
set_gtsummary_theme(theme_gtsummary_compact())

```

```{r}
sa_drugmarket <- drug_market %>%
  dplyr::select(city.factor:link.yn)

ff_glimpse(sa_drugmarket)
```


Summary of the data

```{r drug-market, results="asis", warning=FALSE}

print(dfSummary(sa_drugmarket, 
                varnumbers   = FALSE,
                valid.col    = FALSE,
                graph.magnif = 0.76), 
      max.tbl.height = 300,
      method = "render")
```

### Characteristics of People Who Supply Drugs 

### Sell to Sellers

### Sell to Consumers

### Seller Stock and Sales During COVID-19

## DISCUSSION

### Limitations

## CONCLUSION


\pagebreak

## References

\singlespacing

::: {#refs}
:::

\pagebreak

